Reasoning About a Machine with Local Capabilities

### 2019-02-10 toplas/paper.tex
[TOPLAS] Add discussion on lifting to realistic machine

\section{A Capability Machine with Local Capabilities}
\label{sec:capab-mach-with}
In this paper, we work with a formalization of a capability machine with all the characteristics of real capability machines as well as local capabilities much like CHERI's.
Otherwise, it is kept as simple as possible.
It is inspired by both the M-Machine~\citep{Carter:1994:HSF:195473.195579} and CHERI~\citep{Watson2015Cheri}.
To avoid tedious details irrelevant to the focus of this paper, we assume an infinite address space and unbounded integers.

@@

\itoplassug{We could also include a brief discussion of concurrency and the fact that this probably wouldn't scale to it (but well-bracktedness is not the same in that case anyway.) }
\itoplassug{Reviewer C, esop, wants to know what happens when there are several stack blocks (see tex for comment). We could here emphasize that while we talk about a single stack there could be multiple local read/write-local/execute-capabilities. If we get the stack from an untrusted source, which is the case in the awkward example, then we will use anything that \emph{looks like} the stack as the stack. However, due to the operational semantics for local capabilities all other ``stack blocks'' would have to be stored on the stack or in the register file. As the calling convention dictates that we clear these, the only place the alternative stack blocks will not be cleared is on a part of the stack that we do not have access to. The return capability would basically have to encapsulate this part of memory (perhaps indirectly) which means that from the moment we have been called, we will make sure there is only one stack available until our program returns from the initial call to it.}
%p.10 "to prevent this from happening, we need to make sure the stack capability carries RWLX authority since the system wide assumption then tells us that the adversary cannot have global capabilities to our local stack".  It feels like the "L" capabitility is being overloaded to mean not just "this memory block has write-local authority and can contain local capabilities", but also to mean "this memory block is my stack and no other block is".  What happens when there are several stack blocks, e.g. one per thread?
\emph{Single stack} A single stack is a good choice for the simple capability
machine presented here, because it works well with higher-order functions. An
alternative to a single stack would be to have a separate stack per component.
The trouble with this approach is that, with multiple stacks and local stack
pointers, it is not clear how components would retrieve their stack pointer upon
invocation without compromising safety. A safe approach could be to have stack
pointers stored by a central, trusted stack management component, but it is not
clear how that could scale to large numbers of separate components. Handling
large numbers of components is a requirement if we want to use
capability machines to enforce encapsulation of, for example, every object in an
object-oriented program or every closure in a functional program.
\\\\
\begin{toplas}
\noindent \textbf{Reasoning about capability machine programs}\\
\emph{Limitations} The logical relation defined in
\sectionname~\ref{sec:logical-relation} allows us to reason about capability machine
programs. A limitation w.r.t. previous work is that the logical relation is
tailored towards a specific class of problems.

>>>>

\section{A Capability Machine with Local Capabilities}
\label{sec:capab-mach-with}
In this paper, we work with a formalization of a capability machine with all the characteristics of real capability machines as well as local capabilities much like CHERI's.
Otherwise, it is kept as simple as possible.
It is inspired by both the M-Machine~\citep{Carter:1994:HSF:195473.195579} and CHERI~\citep{Watson2015Cheri}.
To avoid tedious details irrelevant to the focus of this paper, we assume an
infinite address space and unbounded integers (see
\sectionname~\ref{sec:discussion} for more on this).

@@

\itoplassug{We could also include a brief discussion of concurrency and the fact that this probably wouldn't scale to it (but well-bracktedness is not the same in that case anyway.) }
\itoplassug{Reviewer C, esop, wants to know what happens when there are several stack blocks (see tex for comment). We could here emphasize that while we talk about a single stack there could be multiple local read/write-local/execute-capabilities. If we get the stack from an untrusted source, which is the case in the awkward example, then we will use anything that \emph{looks like} the stack as the stack. However, due to the operational semantics for local capabilities all other ``stack blocks'' would have to be stored on the stack or in the register file. As the calling convention dictates that we clear these, the only place the alternative stack blocks will not be cleared is on a part of the stack that we do not have access to. The return capability would basically have to encapsulate this part of memory (perhaps indirectly) which means that from the moment we have been called, we will make sure there is only one stack available until our program returns from the initial call to it.}
%p.10 "to prevent this from happening, we need to make sure the stack capability carries RWLX authority since the system wide assumption then tells us that the adversary cannot have global capabilities to our local stack".  It feels like the "L" capabitility is being overloaded to mean not just "this memory block has write-local authority and can contain local capabilities", but also to mean "this memory block is my stack and no other block is".  What happens when there are several stack blocks, e.g. one per thread?
\emph{Single stack} A single stack is a good choice for the simple capability
machine presented here, because it works well with higher-order functions. An
alternative to a single stack would be to have a separate stack per component.
The trouble with this approach is that, with multiple stacks and local stack
pointers, it is not clear how components would retrieve their stack pointer upon
invocation without compromising safety. A safe approach could be to have stack
pointers stored by a central, trusted stack management component, but it is not
clear how that could scale to large numbers of separate components. Handling
large numbers of components is a requirement if we want to use
capability machines to enforce encapsulation of, for example, every object in an
object-oriented program or every closure in a functional program.
\\\\
\begin{toplas}
  \noindent \textbf{Capability machine formalization}\\
  \emph{Simplifications} Our capability machine formalization assumes unbounded integers and an infinite address space.
  Further, it has a much simpler instruction set than that of a full ISA.
  By making these simplifications, we avoid tedious details but end up with a less realistic machine.
  However, the intend of this work is to gain ground in the formal work necessary to prove properties about low-level assembly programs on a capability machine not to apply it to a full fledged capability machine.

  We do not believe that any of our simplifications are beyond reason.
  If we added unbounded integers, we would have to change the operational
  semantics to take over- and under flows into account. This could be achieved
  by changing the \texttt{plus} and \texttt{minus} instruction to use modulo
  arithmetic.
  If we added a bounded address space, then we would have to take memory out of
  bounds into account. This would require changes to the memory
  operations (\texttt{store} and \texttt{load}) and the step relation.
  The malloc specification would also need to be updated as it would have to
  fail when it runs out of memory.
  Finally, expanding the instruction set to a realistic ISA would add a handful
  of instructions that we would need to reason about.
  All in all, it would add a lot of tedious work to expand this result to a
  realistic machine. The amount of details in the proofs are already at the
  threshold for what should be done with pen and paper, so expanding this work
  to a realistic machine would require mechanized proofs that can take care of
  the tedious details.
\end{toplas}
\\\\
\begin{toplas}
\noindent \textbf{Reasoning about capability machine programs}\\
\emph{Limitations} The logical relation defined in
\sectionname~\ref{sec:logical-relation} allows us to reason about capability machine
programs. A limitation w.r.t. previous work is that the logical relation is
tailored towards a specific class of problems.
